Freezing of a Weeks-Chandler-Anderson Fluid using Classical Density Functional Theory

### 2021-01-12 papers/thesis-kirstie/thesis.tex
edits

Classical Density Functional Theory will be used to find the 
Helmholtz free energies of the fluid and the solid for various 
temperatures and number densities, so that phase diagrams can be 
constructed. 
The key idea behind classical Density Functional Theory (cDFT) is that 
free energy can be written as a functional (a function of a function) of 
a number density profile $n(\vec{r})$ which describes the way the number 
density, or atoms per volume, varies spatially on average. 
\begin{displaymath}{f[n(}\vec{r}{)]}\end{displaymath}
This, coupled with the fact that the free energy of a system tends toward 
a minimum at equilibrium means that the equilibrium free energy of the 
system ${f_{eq}}{[}\rho{(}\vec{r}{)]}$, and its corresponding equilibrium 
number density profile $n_{eq}(\vec r)$, can be found simply by varying 
$n(\vec{r})$ through all possible spatial profiles until the free energy 
is minimized~\cite{MoritaDFT}. 
\begin{displaymath}f[n(\vec r)]\rightarrow f_{eq}[n_{eq}(\vec r)]  \mbox{ as $f$ is minimized} \end{displaymath}
\begin{displaymath}{\mbox{where }  n(\vec{r})=n_{eq}(\vec r)  \mbox{ at equilibrium}}\end{displaymath}
The reason the minimization works is because DFT satisfies the conditions 
of the variational principle. 
%the same mathematical principle used to determine 
%that the shortest distance between two points is a straight line by 
%varying the path between the two points until the distance is minimized.

@@

Rosenfeld also recognized that the Heaviside function can be written as 
a convolution of the weight functions which, for a fluid composed of 
spheres of radius R, is given by

@@

                %&\text{or replacing u notation with r} \nonumber \\

@@

where the weighted densities $n_\alpha$ all depend on the spatially varying 

@@

    n_\alpha(\vec u)=\int n(\vec {r}~')w_\alpha(\vec u-\vec {r}~')~d\vec {r}~'  

@@

Lifting the restriction to the low density limit, Rosenfeld derived an
expression for the excess Helmholtz free energy by noting that the higher
order terms in the virial expansion can all be written in terms of an 
"exclusion volume" (given by Equation~\ref{exclusion-volume}), 
and an "overlap volume" which is the volume of intersection between two 
overlapping spherical regions.
The exclusion volume is the spherical volume about the center of one sphere 
which the center of another sphere can not enter, and is used to describe
terms with $f_{ij}=f(|\vec r_i - \vec r_j|)$ as has been seen. Used in conjunction 
with the overlap volume, the higher order terms in the expansion
in Equation~\ref{eq:Fexcess-simplified}~\cite{rosenfeld1988} can be described. 
%
The resulting expression for the excess Helmholtz free energy with an 
updated version of $\Phi_3$ that was later constructed by Tarazona is given by
\begin{align}
    F_{excess}[n(\vec{r})] &= k_BT\int(\Phi_1(\vec{r})+\Phi_2(\vec{r})+\Phi_3(\vec{r}{)) d}\vec{r} \\
    %\Phi_1 &= -n_{0}ln(1-n_{3})  \\
    %\Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
    %\Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2}
    \Phi_1 &= -n_{0}\ln(1-n_{3}) \\
    \Phi_2 &= \frac{n_{1}n_{2}-\vec{n_{1}}\cdot\vec{n_{2}}}{1-n_{3}} \\
    \Phi_3 &= \frac{{n_2}^3-3n_2\vec{n}_{v2}\cdot\vec{n}_{v2}+\frac{9}{2}[\vec{n}_{v2}\cdot{\overleftrightarrow{n}_{m2}}\cdot{\vec{n}_{v2}}-\operatorname{Tr}({\overleftrightarrow{n}^3_{m2}})]}{24\pi(1-n_3)^2}  
\end{align}
where in addition to the weight functions already defined,  
\begin{align} 
      w_3 &= \Theta(r-R)  \label{scalar-weight3} \\
      w_2 &= \delta(r-R)  \\
      w_1 &= \frac{w_2}{4\pi r}  \\
      w_0 &= \frac{w_2}{4\pi r^2} \label{scalar-weight0}  \\
      \vec {w}_2 &= w_2~\hat r \\
      \vec {w}_1 &= w_1~\hat r
\end{align}
a tensor weight function is given by
\begin{align}\label{eq:tensorweight-methods}
     \overleftrightarrow{w}_{m2}(\vec{r}) = w_2(r)\left(\frac{\vec{r}\vec{r}}{r^2}-\frac{I}{3}\right) 
\end{align}
%The weighted densities $n_0, n_1, n_2, n_3, \vec{n}_1, \vec{n}_2$ and $\overleftrightarrow{n}_{m2}$
%all depend on the spatially varying number density profile $n(\vec r)$.
%\begin{align}%\label{eq:numdenprofile}
	%n_\alpha(\vec{r})=\int_{Volume}{n(\vec{r'})w_\alpha(\vec{r}-\vec{r'})d{\vec{r'}}}
%\end{align}
This result reduces to the same result obtained from Scaled Particle Theory 
for a homogeneous hard-sphere fluid.

>>>>

Classical Density Functional Theory will be used to find the 
Helmholtz free energies of the fluid and the solid for various 
temperatures and number densities, so that phase diagrams can be 
constructed. 
The key idea behind classical Density Functional Theory (cDFT) is that 
free energy can be written as a functional (a function of a function) of 
a number density profile $n(\vec{r})$ which describes the way the number 
density, or atoms per volume, varies spatially on average. 
\begin{displaymath}{f[n(}\vec{r}{)]}\end{displaymath}
This, coupled with the fact that the free energy of a system tends toward 
a minimum at equilibrium means that the equilibrium free energy of the 
system $f_{eq}[n_{eq}(\vec r)]$, and its corresponding equilibrium 
number density profile $n_{eq}(\vec r)$, can be found simply by varying 
$n(\vec{r})$ through all possible spatial profiles until the free energy 
is minimized~\cite{MoritaDFT}. 
\begin{displaymath}f[n(\vec r)]\rightarrow f_{eq}[n_{eq}(\vec r)]  \mbox{ as $f$ is minimized} \end{displaymath}
\begin{displaymath}{\mbox{where }  n(\vec{r})=n_{eq}(\vec r)  \mbox{ at equilibrium}}\end{displaymath}
The reason the minimization works is because DFT satisfies the conditions 
of the variational principle. 
%the same mathematical principle used to determine 
%that the shortest distance between two points is a straight line by 
%varying the path between the two points until the distance is minimized.

@@

where R is either $R_i$ or $R_j$ accordingly. Rosenfeld also recognized that the 
Heaviside function can be written as a convolution of the weight functions 
which, for a fluid composed only of spheres of radius R, is given by

@@

                &\text{Or, replacing u notation with r} \nonumber \\

@@

The weighted densities $n_\alpha$ all depend on the spatially varying 

@@

    n_\alpha(\vec r) &= \int n(\vec {r}~')w_\alpha(\vec r-\vec {r}~')~d\vec {r}~'  \label{weighted_densities}  \\
                     &= n\otimes w_\alpha(\vec r)

@@

Lifting the restriction to the low density limit, Rosenfeld derived an
expression for the excess Helmholtz free energy by noting that the higher
order terms in the virial expansion can all be written in terms of an 
"exclusion volume" (given by Equation~\ref{exclusion-volume}), 
and an "overlap volume" which is the volume of intersection between two 
overlapping spherical regions.
The exclusion volume is the spherical volume about the center of one sphere 
which the center of another sphere can not enter, and is used to describe
terms with $f_{ij}=f(|\vec r_i - \vec r_j|)$ as has been seen. Used in conjunction 
with the overlap volume, the higher order terms in the expansion
in Equation~\ref{eq:Fexcess-simplified}~\cite{rosenfeld1988} can be described. 
%
The resulting expression for the excess Helmholtz free energy with Tarazona's $\Phi_3$
(see Introduction) is given by
\begin{align}
    F_{excess}[n(\vec{r})] &= k_BT\int(\Phi_1(\vec{r})+\Phi_2(\vec{r})+\Phi_3(\vec{r}{)) d}\vec{r} \\
    %\Phi_1 &= -n_{0}ln(1-n_{3})  \\
    %\Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
    %\Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2}
    \Phi_1 &= -n_{0}\ln(1-n_{3}) \\
    \Phi_2 &= \frac{n_{1}n_{2}-\vec{n_{1}}\cdot\vec{n_{2}}}{1-n_{3}} \\
    \Phi_3 &= \frac{{n_2}^3-3n_2\vec{n}_{v2}\cdot\vec{n}_{v2}+\frac{9}{2}[\vec{n}_{v2}\cdot{\overleftrightarrow{n}_{m2}}\cdot{\vec{n}_{v2}}-\operatorname{Tr}({\overleftrightarrow{n}^3_{m2}})]}{24\pi(1-n_3)^2}  
\end{align}
%where in addition to the weight functions already defined,  
where the weighted densities are given by Equation~\ref{weighted_densities} with 
weight functions
\begin{align} 
      w_3 &= \Theta(r-R)  \label{scalar-weight3} \\
      w_2 &= \delta(r-R)  \\
      w_1 &= \frac{w_2}{4\pi r}  \\
      w_0 &= \frac{w_2}{4\pi r^2} \label{scalar-weight0}  \\
      \vec {w}_2 &= w_2~\hat r \\
      \vec {w}_1 &= w_1~\hat r
\end{align}
and a tensor weight function given by
\begin{align}\label{eq:tensorweight-methods}
     \overleftrightarrow{w}_{m2}(\vec{r}) = w_2(r)\left(\frac{\vec{r}\vec{r}}{r^2}-\frac{I}{3}\right) 
\end{align}
%The weighted densities $n_0, n_1, n_2, n_3, \vec{n}_1, \vec{n}_2$ and $\overleftrightarrow{n}_{m2}$
%all depend on the spatially varying number density profile $n(\vec r)$.
%\begin{align}%\label{eq:numdenprofile}
	%n_\alpha(\vec{r})=\int_{Volume}{n(\vec{r'})w_\alpha(\vec{r}-\vec{r'})d{\vec{r'}}}
%\end{align}
This result reduces to the same result obtained from Scaled Particle Theory 
for a homogeneous hard-sphere fluid.
